---
title: "Explaining GAMs"
author: "Nick Giangreco"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install.packages("pacman")
pacman::p_load(data.table,tidyverse,knitr,splines,ggfortify,mgcv)

theme_set(theme_bw(base_size = 16) + theme(text = element_text(face="bold")))
```

# Punchline 

GAM > GLM > Logistic Regression for understanding nonlinear relationships. 

# Introduction

Oftentimes we want to model the relationship between a characteristic and an outcome. In my work, I want to evaluate the characteristic's relationship to the outcome in a way that doesn't bias the type of relationship it naturally has. In this tutorial, I will show how generalized additive models are a flexible yet interpretable way to model relationships in a population and have advantages over traditionally used methods. 

# Outline

* Question

    - Setup the question to be investigated.
   
* Data

    - Display the data to be used to answer the question

* Relationship between outcome and characteristic

    - Show different ways of visualizing the data as well as the relationship between the outcome and characteristic.
   
* Modeling

    - Different modeling strategies to investigate the relationship. Each strategy or method will be generally explained with examples. 
    
    + Logistic Regression

    + Regression Spline

    + Generalized Additive Model
    
    
# Data

```{r data_setup, eval=F,echo=F}

dat <- fread("../data/pediatric_patient_ADEs_atc5th_meddrapt_rinfo_dbid_chid.csv")

report_dat <- 
    dat[,.(safetyreportid,reporter_qualification,sex,age,nichd,weight,receive_date)] %>% 
    unique()

report_dat$D = as.integer(report_dat$safetyreportid %in% dat[atc_concept_id==21604757,safetyreportid,])
report_dat$E = as.integer(report_dat$safetyreportid %in% dat[meddra_concept_id==36919075,safetyreportid])

report_dat %>% 
    fwrite("../data/drug_adverse_event_fda_drug_event_reports.csv")

```

```{r data}

report_dat <- 
    fread("../data/drug_adverse_event_fda_drug_event_reports.csv") %>% 
    .[order(age)]

report_dat[1:5] %>% 
    kable("simple")

```

# Relationship between Paranoia and report characteristics

```{r relationships}

tab <- report_dat[,table(D,E)]
prr <- (tab["1","1"] / ( tab["1","1"] + tab["1","0"] ) ) / (tab["0","1"] / (tab["0","1"] + tab["0","0"]) )
prr

report_dat[E==1,.N,.(D,age)] %>% 
    ggplot(aes(age,N)) +
    geom_point() +
    facet_grid(D~.) +
    ylab("Number of reports with the event") +
    xlab("Age")

report_dat[E==1,.(N = .N),.(D,reporter_qualification)] %>% 
    ggplot(aes(N,reporter_qualification,fill=factor(D))) +
    geom_bar(stat="identity",position="dodge") +
    guides(fill=guide_legend(title="D")) +
    xlab("Number of reports with the event") +
    ylab("")

report_dat[E==1,.(N = .N),.(D,sex)] %>% 
    ggplot(aes(N,sex,fill=factor(D))) +
    geom_bar(stat="identity",position="dodge") +
    guides(fill=guide_legend(title="D")) +
    xlab("Number of reports with the event") +
    ylab("")

report_dat[E==1,.N,.(D,weight)] %>% 
    ggplot(aes(weight,N)) +
    geom_point() +
    scale_y_continuous(trans="log10") +
    facet_grid(D~.) +
    ylab("Number of reports with the event") +
    xlab("Weight")

```

# Modeling

## Logistic Regression

```{r logit}

glm_coef <- function(form="E ~ D - 1",id="id"){
    formula=as.formula(form)
    mod <- glm(formula,data=report_dat,family=binomial(link="logit"))
    summ <- summary(mod)
    dt <- summ$coefficients %>% data.table()
    dt$formula <- form
    dt$term <- summ$coefficients %>% rownames()
    dt$id=id
    dt
}

glm_coefs <- 
    bind_rows(
        glm_coef("E ~ D",id='model1'),
        glm_coef("E ~ age",id='model2'),
        glm_coef("E ~ D + age",id='model3')
        )

glm_coefs %>% 
    ggplot(aes(formula,Estimate,color=term,group=term)) +
    geom_point() +
    geom_line()

age_coef <- glm_coef("E ~ age",id='model2')[term=="age",Estimate]
age_coef2 <- glm_coef("E ~ age + D",id='model2')[term=="age",Estimate]

data.table(Age = seq(0,21,1),`E ~ age` = age_coef*seq(0,21,1),`E ~ age + D` = age_coef2*seq(0,21,1)) %>% 
    pivot_longer(cols=c("E ~ age","E ~ age + D")) %>% 
    ggplot(aes(Age,value,color=name)) +
    geom_point() +
    geom_line() +
    guides(color=guide_legend(title="Model formula")) +
    xlab("Age") +
    ylab("Risk of Paranoia")

```

## Polynomial Regression

`r age_grid <- seq(0,21)`

```{r poly}
# http://www.science.smith.edu/~jcrouser/SDS293/labs/lab12-r.html
fit_plot_polynomial <- function(degree=3){
    fit = glm(E ~ poly(age, degree), data = report_dat)
    preds = predict(fit, newdata = list(age = age_grid), se = TRUE)
    pfit = exp(preds$fit) / (1+exp(preds$fit))
    se_bands_logit = cbind("upper" = preds$fit+2*preds$se.fit, 
                           "lower" = preds$fit-2*preds$se.fit)
    se_bands = exp(se_bands_logit) / (1+exp(se_bands_logit))
    
    E = report_dat[E==1]
    
    notE = report_dat[E==0]
    
    coefs = sapply(coef(fit)[2:(degree+1)],function(x){round(x,2)})
    formula = 
        paste0("E ~ ",paste0(sapply(1:degree,function(x){paste0(coefs[x]," * age^",x)}),collapse=" + "))
    
    ggplot() +
      geom_rug(data = notE, aes(x = jitter(age), y = max(pfit)), sides = "b", alpha = 0.1) +
      geom_rug(data = E, aes(x = jitter(age), y = min(pfit)), sides = "t", alpha = 0.1) +
      geom_line(aes(x = age_grid, y = pfit), color = "#0000FF") +
      geom_ribbon(aes(x = age_grid, 
                      ymin = se_bands[,"lower"], 
                      ymax = se_bands[,"upper"]), 
                  alpha = 0.3) +
      labs(title = paste0(formula,"\nAIC=",round(fit$aic,0)),
           x = "Age",
           y = "P(E)")
}

fit_plot_polynomial(1)
fit_plot_polynomial(2)
fit_plot_polynomial(3)
fit_plot_polynomial(4)

```

## Regression splines

```{r splines_broken}

# Simon Wood's Intro to GAMs textbook chapter 4

tf <- function(x,xj,j){
    ## generate the jth tent function from set defined by knots xj
    dj <- xj*0;dj[j] <- 1
    ## linearly interpolates between xj and dj taking place at x
    approx(xj,dj,x)$y
}

tf.X <- function(x,xj){
    ## tent function basis matrix given data X
    ## and knot sequence xk
    nk <- length(xj); n <- length(x)
    X <- matrix(NA,n,nk)
    for(j in 1:nk) X[,j] <- tf(x,xj,j)
    X
}

knots = c(0,1,2,6,11,17,19)
Xp <- tf.X(age_grid,knots)

Xp_dt = data.table(Xp)
colnames(Xp_dt) <- knots %>% as.character()
Xp_dt$age <- age_grid
Xp_dt_melt <- 
    Xp_dt %>% 
    melt(id.vars="age",
         variable.name="knot")

Xp_dt_melt %>% 
    ggplot(aes(age,value,color=knot)) +
    geom_point() +
    geom_line() +
    facet_grid(knot~.)
```


```{r splines}
# https://stats.stackexchange.com/questions/29345/visualizing-a-spline-basis/29346
x <- report_dat$age
spl <- ns(x,knots=knots)

#sapply(1:ncol(spl),function(i){plot(x,spl[,i],ylab=paste0("Basis ",i),xlab="Age")})

autoplot(spl)

```

## Generalized additive models

<!-- https://rdrr.io/cran/mgcv/man/smooth.terms.html --->

```{r glm_splines}

mod = mgcv::bam(E ~ s(age) + D,
                data=report_dat,
                family=binomial(link="logit"),
                method = "fREML")

mod 

plot(mod)

mod = mgcv::bam(E ~ s(age,bs="cs") + D,
                data=report_dat,
                family=binomial(link="logit"),
                method = "fREML")

mod

plot(mod)

mod = mgcv::bam(E ~ s(age,bs="ps") + D,
                data=report_dat,
                family=binomial(link="logit"),
                method = "fREML")

mod

plot(mod)

```

